---
title: "Parquet & Feather 2/3: Security Telemetry Benchmark"
authors:
  - dispanser
  - mavam
date: 2022-10-15
tags: [arrow, parquet, feather, quarto, r]
---

How does Apache [Parquet][parquet] compare to [Feather][feather] for storing
structured security data? In this blog post, we answer this question.

[parquet]: https://parquet.apache.org/
[feather]: https://arrow.apache.org/docs/python/feather.html

<!--truncate-->

:::note Parquet & Feather: 2/3
This is blog post is part of a 3-piece series on Parquet and Feather.

1. Enabling Open Investigations
2. Security Telemetry Benchmark
3. TBD

:::

In the previous blog post, we explained why Parquet and Feather are great
building blocks for modern investigations. We now want to take a closer look at
how they actually perform with respect to the following metrics:

- **Size**: how much space does typical security telemetry occupy?
- **Write throughput**: how fast can we write out to a store?

Parquet and Feather have different goals. While Parquet is an on-disk format
that optimizes for size, Feather is just a thin layer around the native Arrow
in-memory representation. This puts them at different points in the spectrum of
throughput and latency. To better understand this spectrum for real-world data,
we instrumented the write path of VAST, which consists roughly of the following
steps:

1. Parse the input
2. Convert it into Arrow record batches
3. Ship Arrow record batches to a VAST server
4. Write Arrow out into a Parquet or Feather store

Since steps (1–3) are the same for both stores, we ignore them in the following
analysis and solely zoom in on (4). For our measurements, we collect the
following data points:

- **Store**: the type of store plugin used in the measurement, i.e., `parquet`
  or `feather`.

- **Construction time**: the time it takes to convert Arrow record batches into
  Parquet or Feather. We fenced the corresponding code blocks and computed the
  difference microseconds.

- **Input size**: the number of bytes that the to-be-converted record batches
  consume.

- **Output size**: the number of bytes that the store file takes up.

- **Number of events**: the total number of events in all input record batches

- **Number of record batches**

- **Schema**: the name of the schema; there exists one store file per schema

## Dataset

TODO: Describe that dataset

```{r}
#| code-fold: true
#| label: prepare-data
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(tidyr)

theme_set(theme_minimal())

data <- read.csv("data.csv") |>
  mutate(duration = dnanoseconds(as.numeric(gsub("ns", "", duration))))

sizes <- read.csv("sizes.csv") |>
  mutate(store_type = "original")

# Normalize store sizes by number of events/store.
normalized <- data |>
  mutate(duration_normalized = duration / num_events,
         bytes_memory_normalized = bytes_memory / num_events,
         bytes_storage_normalized = bytes_in_storage / num_events,
         bytes_ratio = bytes_in_storage / bytes_memory)

# Compute average over measurements.
aggregated <- normalized |>
  group_by(store_type, schema) |>
  summarize(duration = mean(duration_normalized),
            memory = mean(bytes_memory_normalized),
            storage = mean(bytes_storage_normalized))

# Treat in-memory measuremts as just another storage type.
memory <- aggregated |>
  filter(store_type == "feather") |>
  mutate(store_type = "memory") |>
  select(-storage) |>
  rename(bytes = memory)

# Unite with rest of data.
unified <- aggregated |>
  select(-memory) |>
  rename(bytes = storage) |>
  bind_rows(memory, sizes)
```

We have a total of `r length(unique(data$schema))` unique schemas:

```{r}
#| label: schemas
unique(data$schema)
```

The schemas belong to three data types: Zeek, Suricata, and PCAP. These are all
data sources VAST can ingest natively. Zeek and Suricata are ASCII, whereas PCAP
is a binary format representing a stream of network packets.

Our dataset has multiple different schemas, each of which has a different number
of events.

```{r}
#| code-fold: true
#| label: number-of-events
schemas <- normalized |>
  filter(store_type == "feather") |>
  group_by(schema) |>
  summarize(n = median(num_events))

schemas |>
  separate(schema, c("module", "type"), remove = FALSE) |>
  ggplot(aes(x = reorder(schema, -n), y = n, fill = module)) +
    geom_bar(stat = "identity") +
    guides(fill = "none") +
    scale_y_log10(labels = scales::label_comma()) +
    labs(x = "Schema", y = "Number of Events", color = "Module") +
    theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1))
```

Zeek connection logs lead the pack, followed by PCAP packets.

## Size

To better understand the difference between Parquet and Feather, we now take a
look at them right next to each other. In addition to Feather and Parquet, we
use three other types of "stores" for the analysis to faciliate comparison:

1. **Original**: the size of the input prior it entered VAST, e.g., the raw JSON or
   a PCAP file.

2. **Segment**: our custom store prior to having switched to Feather. We omit
   this reference point often, as it is nearly equivalent to Feather.

3. **Memory**: the size of the data in memory, measured as the sum of Arrow
   buffers that make up the table slice.

Let's kick of the analysis by getting a better understanding at the distribution
of size. 

```{r}
#| code-fold: true
#| label: plot-schema-distribution
unified |>
  ggplot(aes(x = bytes, color = store_type)) +
  geom_step(stat = "ecdf") +
  scale_x_log10(labels = scales::label_bytes(units = "auto_si")) +
  labs(x = "Bytes/Event", y = "ECDF", color = "Store")
```

```{r}
#| code-fold: true
#| label: plot-parquet-vs-feather-size
library(ggrepel)

# Plot: Scatterplot of Feather vs. Parquet for Feather events < 100 bytes
unified |>
  filter(store_type != "segment") |>
  pivot_wider(names_from = store_type, values_from = bytes, id_cols = schema) |>
  filter(feather < 100) |>
  separate(schema, c("module", "type"), remove = FALSE) |>
  ggplot(aes(x = parquet, y = feather, color = module, size = original)) +
    geom_abline(intercept = 0, slope = 1, color = "grey") +
    geom_point(alpha = 0.6) +
    geom_text_repel(aes(label = schema),
              color = "grey",
              size = 2, # font size
              box.padding = 0.5,
              min.segment.length = 0, # draw all line segments
              max.overlaps = Inf,
              segment.color = "grey",
              segment.alpha = 0.5) +
    xlim(0, 100) +
    ylim(0, 100) +
    scale_size(range = c(0, 10)) +
    guides(color = "none") +
    labs(size = "Original size", x = "Bytes (Parquet)", y = "Bytes (Feather)")
```

Let's pick

```{r}
#| code-fold: true
#| label: plot-zeek-suricata
unified |>
  filter(schema %in% c("zeek.conn", "suricata.snmp")) |>
  ggplot(aes(x = reorder(store_type, -bytes), y = bytes, fill = store_type)) +
    geom_bar(stat = "identity") +
    guides(fill = "none") +
    labs(x = "Store", y = "Median Bytes/Event") +
    scale_y_continuous(labels = scales::label_bytes(units = "auto_si")) +
    facet_grid(~ schema)
```

## Write Throughput

With write throughput, we mean the time it takes to transform Arrow Record
Batches into Parquet and Feather format.


```{r}
#| code-fold: true
#| label: compute-bytes-per-event
#aggregated |>
#  ggplot(aes(x = reorder(schema, -bytes), y = bytes, fill = store_type)) +
#    geom_bar(stat = "identity", position = "dodge") +
#    expand_limits(y = 1) +
#    xlab("Schema") +
#    ylab("Bytes") +
#    scale_y_continuous(labels = scales::label_bytes(units = "auto_si")) +
#    theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1))
```

```{r}
#| code-fold: true
#| label: plot-parquet-vs-feather-duration
aggregated |>
  filter(store_type != "segment") |>
  mutate(duration = dseconds(duration)) |>
  pivot_wider(names_from = store_type, values_from = duration, id_cols = schema) |>
  separate(schema, c("module", "type"), remove = FALSE) |>
  ggplot(aes(x = parquet, y = feather, color = module)) +
    geom_abline(intercept = 0, slope = 1, color = "grey") +
    geom_point(size = 4, alpha = 0.6) +
    geom_text_repel(aes(label = schema),
              color = "grey",
              size = 2, # font size
              box.padding = 0.3,
              min.segment.length = 0, # draw all line segments
              max.overlaps = Inf,
              segment.color = "grey",
              segment.alpha = 0.5) +
    xlim(0, dmicroseconds(50)) +
    ylim(0, dmicroseconds(50)) +
    scale_size(range = c(0, 10)) +
    guides(color = "none") +
    labs(x = "Bytes (Parquet)", y = "Bytes (Feather)")
```
