---
title: "Parquet & Feather: Security Telemetry Benchmark"
authors:
  - dispanser
  - mavam
date: 2022-10-15
tags: [benchmark, arrow, parquet, feather, quarto, r]
---

How does Apache [Parquet][parquet] compare to [Feather][feather] for storing
structured security data? In this blog post, we answer this question.

[parquet]: https://parquet.apache.org/
[feather]: https://arrow.apache.org/docs/python/feather.html

<!--truncate-->

:::info Parquet & Feather: 2/3
This is blog post is part of a 3-piece series on Parquet and Feather.

1. [Enabling Open Investigations][parquet-and-feather-1]
2. Security Telemetry Benchmark (this blog post)
3. TBD

[parquet-and-feather-1]: /blog/parquet-and-feather-enabling-open-investigations/
:::

In the [previous blog post][parquet-and-feather-1], we explained why Parquet and
Feather are great building blocks for modern investigations. We now want to take
a closer look at how they actually perform with respect to the following
metrics:

- **Size**: how much space does typical security telemetry occupy?
- **Write throughput**: how fast can we write out to a store?

Parquet and Feather have different goals. While Parquet is an on-disk format
that optimizes for size, Feather is just a thin layer around the native Arrow
in-memory representation. This puts them at different points in the spectrum of
throughput and latency. To better understand this spectrum for real-world data,
we instrumented the write path of VAST, which consists roughly of the following
steps:

1. Parse the input
2. Convert it into Arrow record batches
3. Ship Arrow record batches to a VAST server
4. Write Arrow out into a Parquet or Feather store

Since steps (1–3) are the same for both stores, we ignore them in the following
analysis and solely zoom in on (4). For our measurements, we collect the
following data points:

- **Store**: the type of store plugin used in the measurement, i.e., `parquet`
  or `feather`.
- **Construction time**: the time it takes to convert Arrow record batches into
  Parquet or Feather. We fenced the corresponding code blocks and computed the
  difference microseconds.
- **Input size**: the number of bytes that the to-be-converted record batches
  consume.
- **Output size**: the number of bytes that the store file takes up.
- **Number of events**: the total number of events in all input record batches
- **Number of record batches**
- **Schema**: the name of the schema; there exists one store file per schema

## Dataset

:::caution TODO
Describe the dataset.
:::

```{r}
#| code-fold: true
#| label: prepare-data
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(tidyr)

theme_set(theme_minimal())

data <- read.csv("data.csv") |>
  rename(store = store_type) |>
  mutate(duration = dnanoseconds(as.numeric(gsub("ns", "", duration))))

sizes <- read.csv("sizes.csv") |>
  mutate(store = "original")

# Normalize store sizes by number of events/store.
normalized <- data |>
  mutate(duration_normalized = duration / num_events,
         bytes_memory_normalized = bytes_memory / num_events,
         bytes_storage_normalized = bytes_in_storage / num_events,
         bytes_ratio = bytes_in_storage / bytes_memory)

# Compute average over measurements.
aggregated <- normalized |>
  group_by(store, schema, zstd.level) |>
  summarize(duration = mean(duration_normalized),
            memory = mean(bytes_memory_normalized),
            storage = mean(bytes_storage_normalized))

# Treat in-memory measuremts as just another storage type.
memory <- aggregated |>
  filter(store == "feather") |>
  mutate(store = "memory") |>
  select(-storage) |>
  rename(bytes = memory)

# Unite with rest of data.
unified <- aggregated |>
  select(-memory) |>
  rename(bytes = storage) |>
  bind_rows(memory, sizes)
```

### Batching

When working with security event data, we see highly variable data shapes. Some
event types completely cominate a dataset, and the long tail is large.

On the inside, a store is just a concatenation of Arrow record batches. This
means the "batching shape" of the store is defined by two factors: number of
events per batch and total number of batches. Let's plot these against each
other to get a feel of the data:

```{r}
#| code-fold: true
#| label: events
no_segment <- data |>
  filter(store != "segment")

no_segment |>
  ggplot(aes(x = num_events, y = num_slices, size = bytes_in_storage, color = store)) +
    geom_point(alpha = 0.3) +
    scale_size(range = c(2, 10), labels = scales::label_bytes(units = "auto_si")) +
    scale_x_log10(labels = scales::label_comma()) +
    labs(color = "Store", size = "Size", x = "Events", y = "Batches")
```

Each point in the above plot represents one store file. The x-axis (log-scaled)
represents the number of events in the store and the y-axis the number of
batches. The point diameter represents the size of the store file on disk. The
color denotes the type of store: Parquet or Feather. We omit our Segment store,
as it's almost identical to Feather.

```{r}
#| echo: false
top <- no_segment |> arrange(num_events) |> tail(1)
bot <- no_segment |> arrange(num_events) |> head(1)
```

Our dataset has everything from a few dozons to millions of events per store.
The event "`r top |> pull(schema)`" contains a total of
`r top |> pull(num_events) |> label_comma()()` events, whereas the event
"`r bot |> pull(schema)`" has only `r bot |> pull(num_events)`. Parquet files
also have more batches compared to Feather.

:::caution TODO
Thomas, why? :-)
:::

This high variance in number of events per store is suboptimal, because every
partition (that contains the store) has a unique entry and indexes in the
catalog. This causes fragmentation. Ideally, we would like to create a uniform
number of events per partition. Thankfully, VAST users do not have to worry
about this churn. As of [v2.3](/blog/vast-v2.3), we have automatic rebuilding in
place, which merges underfull partitions to reduce pressure on the central
catalog.

How does the number of events relates to their size?

```{r}
#| code-fold: true
#| label: log-log-event-memory
no_segment |>
  ggplot(aes(x = num_events, y = bytes_memory, color = store)) +
    geom_smooth(method = "lm", se = FALSE, size = 0.5) +
    geom_point(alpha = 0.5) +
    guides(color = "none") +
    scale_y_log10(labels = scales::label_bytes(units = "auto_si")) +
    scale_x_log10(labels = scales::label_comma()) +
    labs(color = "Store", x = "Events", y = "Bytes (in memory)") +
    facet_grid(~ store)
```

The above [log-log plot](https://en.wikipedia.org/wiki/Log%E2%80%93log_plot)
includes a fitted linear model. Except for the tails, we can see that the points
align on a straight line. This means that there exists a power-function
relationship between the number of events and the store size. The slope of the
line is the exponent and precisely captures this relationship.

For the remainder of the analysis, we look at the normalized event size, by
dividing the store size by the number of contained events.

### Schemas

We have a total of `r length(unique(data$schema))` unique schemas:

```{r}
#| label: schemas
#| echo: false
unique(data$schema)
```

The schemas belong to three data types: Zeek, Suricata, and PCAP. These are all
data sources VAST can ingest natively. Zeek and Suricata are text, whereas PCAP
is a binary format representing a stream of network packets.

Our dataset has multiple different schemas, each of which has a different number
of events:

```{r}
#| code-fold: true
#| label: number-of-events-by-schema
schemas <- normalized |>
  filter(store == "feather") |>
  group_by(schema) |>
  summarize(n = median(num_events),
            bytes_memory = median(bytes_memory))

schemas |>
  separate(schema, c("module", "type"), remove = FALSE) |>
  ggplot(aes(x = reorder(schema, -n), y = n, fill = module)) +
    geom_bar(stat = "identity") +
    guides(fill = "none") +
    scale_y_log10(labels = scales::label_comma()) +
    labs(x = "Schema", y = "Number of Events", color = "Module") +
    theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1))
```

Zeek connection logs lead the pack, followed by PCAP packets. Some more
voluminous events represent the long tail, e.g., `suricata.ikev2`. What this
shows is basically that we have a diverse dataset. We already saw the
relationship between the number of events and their (in-memory) size on the
log-log plot earlier, but if we exchange number of events with their size, we
can observe almost the same order:

```{r}
#| code-fold: true
#| label: event-size-by-schema
schemas |>
  separate(schema, c("module", "type"), remove = FALSE) |>
  ggplot(aes(x = reorder(schema, -n), y = bytes_memory, fill = module)) +
    geom_bar(stat = "identity") +
    guides(fill = "none") +
    scale_y_log10(labels = scales::label_bytes(units = "auto_si")) +
    labs(x = "Schema", y = "Bytes (in-memory)", color = "Module") +
    theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1))
```

## Size

To better understand the difference between Parquet and Feather, we now take a
look at them right next to each other. In addition to Feather and Parquet, we
use three other types of "stores" for the analysis to faciliate comparison:

1. **Original**: the size of the input prior it entered VAST, e.g., the raw JSON or
   a PCAP file.

2. **Segment**: our custom store prior to having switched to Feather. We omit
   this reference point often, as it is nearly equivalent to Feather.

3. **Memory**: the size of the data in memory, measured as the sum of Arrow
   buffers that make up the table slice.

Let's kick of the analysis by getting a better understanding at the distribution
of size.

```{r}
#| code-fold: true
#| label: plot-schema-distribution
unified |>
  ggplot(aes(x = bytes, color = store)) +
  geom_step(stat = "ecdf") +
  scale_x_log10(labels = scales::label_bytes(units = "auto_si")) +
  labs(x = "Bytes/Event", y = "ECDF", color = "Store")
```

In the above
[ECDF](https://en.wikipedia.org/wiki/Empirical_distribution_function) plot, we
can see that Parquet stores on the left (in blue) have the smallest size,
followed by Feather (and Segment), and then their corresponding in-memory and
original representation. A good way to read this plot is by locating the median,
which is at y = 0.5, and then find the corresponding value on the x-axis. Here
are the medians:

```{r}
#| echo: false
library(knitr)

medians <- unified |>
  group_by(store) |>
  summarize(median = median(bytes)) |>
  arrange(median)

median_parquet <- medians |> filter(store == "parquet") |> pull(median)
median_feather <- medians |> filter(store == "feather") |> pull(median)
median_original <- medians |> filter(store == "original") |> pull(median)

medians |> kable()
```

This makes it clear from a numbers perspective: there's a factor of
`r round(median_feather / median_parquet)` difference between Parquet and
Feather. Compared to the original data, we're at factors
`r round(median_original / median_parquet)` and
`r round(median_original / median_feather)` for Parquet and Feather,
respectively.

We have to remember that there Zstd compression at play. That said, both are
enabled for Parquet and Feather.

:::caution TODO
TODO: elaborate
:::

If we just compare Parquet and Feather for a given schema, we see the difference
between the two more clearly:

```{r}
#| code-fold: true
#| label: plot-parquet-vs-feather-size-lt100
library(ggrepel)

parquet_vs_feather <- unified |>
  filter(store != "segment") |>
  pivot_wider(names_from = store, values_from = bytes, id_cols = schema)

plot_parquet_vs_feather = function(data) {
  data |>
    separate(schema, c("module", "type"), remove = FALSE) |>
    ggplot(aes(x = parquet, y = feather, color = module, size = original)) +
      geom_abline(intercept = 0, slope = 1, color = "grey") +
      geom_point(alpha = 0.6) +
      geom_text_repel(aes(label = schema),
                color = "grey",
                size = 3, # font size
                box.padding = 0.5,
                min.segment.length = 0, # draw all line segments
                max.overlaps = Inf,
                segment.color = "grey",
                segment.alpha = 0.5) +
      scale_size(range = c(0, 10)) +
      guides(color = "none", size = "none") +
      labs(size = "Original size", x = "Bytes (Parquet)", y = "Bytes (Feather)")
}

parquet_vs_feather |>
  filter(feather < 100) |>
  plot_parquet_vs_feather() +
    theme(aspect.ratio = 1)
```

```{r}
#| code-fold: true
#| label: plot-parquet-vs-feather-size-ge100
parquet_vs_feather |>
  filter(feather >= 100) |>
  plot_parquet_vs_feather() +
    coord_fixed(ratio = 2/5)
```

Every point in the above scatterplots represents a median size of a store for a
given schema. The straight line represents the identity function. Points closer
to the identity have less of a difference between the two formats, whereas
points further away have a stark difference. The first plot has all schemas
where the size per event is less than 100 bytes for the Feather store, and the
second plot contains the remainder. (We used two separate plots, where the first
one has a 1:1 aspect ratio and the second 2:5.)

The region above the identity line is where Feather is larger than Parquet. All
points fall in there. But there are some wide differences in the delta:

```{r}
#| code-fold: true
#| label: plot-zeek-suricata
zeek_conn <- unified |>
  filter(store %in% c("parquet", "feather") & schema == "zeek.conn") |>
  pull(bytes)

suricata_snmp <- unified |>
  filter(store %in% c("parquet", "feather") & schema == "suricata.snmp") |>
  pull(bytes)

suricata_ikev2 <- unified |>
  filter(store %in% c("parquet", "feather") & schema == "suricata.ikev2") |>
  pull(bytes)

unified |>
  filter(schema %in% c("zeek.conn", "suricata.snmp", "suricata.ikev2")) |>
  ggplot(aes(x = reorder(store, -bytes), y = bytes, fill = store)) +
    geom_bar(stat = "identity") +
    guides(fill = "none") +
    labs(x = "Store", y = "Median Bytes/Event") +
    scale_y_continuous(labels = scales::label_bytes(units = "auto_si")) +
    theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1)) +
    facet_wrap(~ schema, scales = "free")
```

For `zeek.conn`, the percentage that Parquet takes compared to Feather is
`r scales::label_percent()(zeek_conn[2] / zeek_conn[1])`, whereas it is
`r scales::label_percent()(suricata_snmp[2] / suricata_snmp[1])` for
`suricata.snmp`.

:::caution TODO
Thomas, we need to look at suricata.ikev2. It's really weird to see the
blow-up.
:::

We finally look at the fraction of space Parquet takes compared to Feather on a
per schema basis:

```{r}
#| code-fold: true
#| label: plot-parquet-divided-by-feather
library(tibble)

perc <- parquet_vs_feather |>
  transmute(schema,
            perc = scales::label_percent(accuracy = 1)(parquet / feather)) |>
  column_to_rownames("schema")

parquet_vs_feather |>
  ggplot(aes(x = reorder(schema, -parquet / feather),
             y = parquet / feather,
             fill = memory)) +
    geom_bar(stat = "identity") +
    labs(x = "Schema", y = "Parquet / Feather (%)", fill = "Bytes\nin Memory") +
    scale_y_continuous(labels = scales::label_percent()) +
    theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1))
```

Overall, we see that `pcap.packet` has almost no improvement with
`r perc["pcap.packet",]`. Most likely because the bulk volume of the packet is
not the structured header, and therefore compresses equally well via Zstd.
Conversely, `zeek.ssl` boils down to `r perc["zeek.ssl",]` of its Feather
equivalent. TODO: figure out why.


## Write Throughput

With write throughput, we mean the time it takes to transform Arrow Record
Batches into Parquet and Feather format.


```{r}
#| code-fold: true
#| label: compute-bytes-per-event
#aggregated |>
#  ggplot(aes(x = reorder(schema, -bytes), y = bytes, fill = store)) +
#    geom_bar(stat = "identity", position = "dodge") +
#    expand_limits(y = 1) +
#    xlab("Schema") +
#    ylab("Bytes") +
#    scale_y_continuous(labels = scales::label_bytes(units = "auto_si")) +
#    theme(axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1))
```

```{r}
#| code-fold: true
#| label: plot-parquet-vs-feather-duration
aggregated |>
  filter(store != "segment") |>
  mutate(duration = dseconds(duration)) |>
  pivot_wider(names_from = store, values_from = duration, id_cols = schema) |>
  separate(schema, c("module", "type"), remove = FALSE) |>
  ggplot(aes(x = parquet, y = feather, color = module)) +
    geom_abline(intercept = 0, slope = 1, color = "grey") +
    geom_point(size = 4, alpha = 0.6) +
    geom_text_repel(aes(label = schema),
              color = "grey",
              size = 2, # font size
              box.padding = 0.3,
              min.segment.length = 0, # draw all line segments
              max.overlaps = Inf,
              segment.color = "grey",
              segment.alpha = 0.5) +
    scale_size(range = c(0, 10)) +
    scale_x_continuous(limits = c(0, dmicroseconds(50)),
                       labels = scales::label_number(scale = 1e6, suffix = "us")) +
    scale_y_continuous(limits = c(0, dmicroseconds(50)),
                       labels = scales::label_number(scale = 1e6, suffix = "us")) +
    guides(color = "none") +
    labs(x = "Duration (Parquet)", y = "Duration (Feather)")
```

:::caution TODO
These experiments need to be redone. It looks like the runtime is completely
dominated by Zstd compression. The expectation would be that all the points are
very far away from the identity line in the bottom right region
:::
